Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-01-14 paper/main.tex

We observed a similar set of results when we restricted our estimates of
participants' knowledge for questions about each lecture to consider only
their performance on other questions about the \textit{same} lecture.
Specifically, for Quiz~1, participants' knowledge of \textit{Four Fundamental
Forces}-related questions, estimated from their performance on other
\textit{Four Fundamental Forces}-related questions, was predictive of their
ability to answer those questions correctly ($OR = 15.934,\ 95\%\
\textnormal{CI} = [5.173,\ 38.005],\ \lambda_{LR} = 40.971,\ p = 0.001$). The
same was true of participants' estimated knowledge for \textit{Birth of
Stars}-related questions based on their performance on other \textit{Birth of
Stars}-related questions ($OR = 9.775,\ 95\%\ \textnormal{CI} = [2.93,\
25.08],\ \lambda_{LR} = 13.924,\ p = 0.001$). These within-lecture knowledge
estimates also predicted success on questions about both lectures when we
computed them analogously for Quiz~2 (\textit{Four Fundamental Forces}: $OR =
35.126,\ 95\%\ \textnormal{CI} = [5.113,\ 123.868],\ \lambda_{LR} = 32.251,\
p < 0.001$; \textit{Birth of Stars}: $OR = 4.717,\ 95\%\ \textnormal{CI} =
[2.021,\ 9.844],\ \lambda_{LR} = 16.788,\ p < 0.001$). For Quiz~3, we found that
within-lecture knowledge estimates predicted participants' success on
\textit{Birth of Stars}-related questions ($OR = 16.902,\ 95\%\ \textnormal{CI}
= [3.353,\ 53.265],\ \lambda_{LR} = 23.233,\ p < 0.001$) but not on \textit{Four
Fundamental Forces}-related questions ($OR = 2.485,\ 95\%\ \textnormal{CI} =
[0.724,\ 8.366],\ \lambda_{LR} = 1.984,\ p = 0.170$). This may indicate that the
within-lecture knowledge estimates are susceptible to ceiling effects in
participants' quiz performance. On Quiz~3, after viewing both lectures, no
participant answered more than three \textit{Four Fundamental Forces}-related
questions incorrectly, and all but five participants (out of 50) answered two or
fewer questions incorrectly. (This was the only subset of questions about either
lecture, across all three quizzes, for which this was true.) Because of this,
when we held out one incorrectly answered 
\textit{Four Fundamental Forces}-related question from a given participant's 
Quiz~3 responses and estimated their knowledge at its embedding coordinate using 
the remaining \textit{Four Fundamental Forces}-related questions they answered, 
for 90\% of participants, that estimate leveraged information about at most a 
single other question they were \textit{not} able to correctly answer. This
broad homogeneity in participants' success on questions used to estimate their 
knowledge may have hurt our ability to accurately characterize the specific (and 
by Quiz~3, relatively few) aspects of the lecture content they did \textit{not} 
know about. Taken together, these results suggest that our knowledge estimates 
can reliably distinguish between questions about different content covered by a 
single lecture, provided there is sufficient diversity in participants' quiz 
responses to extract meaningful information about both what they know and what 
they do not know.

Finally, when we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture, we observed a somewhat different pattern of results.
Here we found that before viewing either lecture (i.e., on Quiz~1),
participants' abilities to answer \textit{Four Fundamental Forces}-related
questions could not be predicted from their responses to \textit{Birth of
Stars}-related questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\
9.088],\ \lambda_{LR} = 0.712,\ p = 0.404$), nor could their abilities to
answer \textit{Birth of Stars}-related questions be predicted from their
responses to \textit{Four Fundamental Forces}-related questions ($OR = 1.522,\
95\%\ \textnormal{CI} = [0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$).
We similarly found that participants' success on questions about either lecture
could not be predicted given their responses to questions about the other
lecture after viewing \textit{Four Fundamental Forces} but before viewing \textit{Birth of Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces}
questions given \textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\
\textnormal{CI} = [0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 2.199,\ 95\%\ \textnormal{CI} = [0.711,\ 5.623],\
\lambda_{LR} = 2.304,\ p = 0.141$). Only after viewing \textit{both} lectures
(i.e., on Quiz~3) did these across-lecture knowledge estimates reliably predict
participants' success on individual quiz questions (\textit{Four Fundamental
Forces} questions given \textit{Birth of Stars} questions: $OR = 11.294,\
95\%\ \textnormal{CI} = [1.375,\ 47.744],\ \lambda_{LR} = 10.396,\ p < 0.001$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 7.302,\ 95\%\ \textnormal{CI} = [1.077,\ 44.879],\
\lambda_{LR} = 4.708,\ p = 0.038$). Taken together, these results suggest that
our knowledge estimates can be used to predict participants' success across
content areas once they have received some training on both the content about
which their knowledge is estimated and the content used to construct these
estimates.

@@

We view our work as making several contributions to the study of how people
acquire conceptual knowledge. First, from a methodological standpoint, our
modeling framework provides a systematic means of mapping out and
characterizing knowledge in maps that have infinite (arbitrarily many) numbers
of coordinates, and of ``filling out'' those maps using relatively small
numbers of multiple choice quiz questions. Our experimental finding that we can
use these maps to predict responses to held-out questions has several
psychological implications as well. For example, concepts that are assigned to
nearby coordinates by the text embedding model also appear to be ``known to a
similar extent'' (as reflected by participants' responses to held-out
questions; Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. How participants' knowledge falls off with spatial distance is
captured by the knowledge maps we infer from their quiz responses
(e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words,
our study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

@@

To generate our estimates, we placed a set of 39 radial basis functions (RBFs)
throughout the embedding space, centered on the 2D projections for each
question (i.e., we included one RBF for each question). At coordinate $x$, the
value of an RBF centered on a question's coordinate $\mu$, is given by:
\begin{equation}
    \mathrm{RBF}(x, \mu, \lambda) = \exp\left\{-\frac{||x - \mu||^2}{\lambda}\right\}.
    \label{eqn:rbf}
\end{equation}
The $\lambda$ term in the RBF equation controls the ``smoothness'' of the
function, where larger values of $\lambda$ result in smoother maps. In our
implementation we used $\lambda = 50$.  Next, we estimated the ``knowledge''
at each coordinate, $x$, using:
\begin{equation}
    \hat{k}(x) = \frac{\sum_{i \in \mathrm{correct}} \mathrm{RBF}(x, q_i, \lambda)}{\sum_{j = 1}^N \mathrm{RBF}(x, q_j, \lambda)}.
    \label{eqn:rbf-knowledge}
\end{equation}
Intuitively, Equation~\ref{eqn:rbf-knowledge} computes the weighted proportion of
correctly answered questions, where the weights are given by how nearby (in the 2D space)
each question is to the $x$.  We also defined \textit{learning maps} as the coordinate-by-coordinate
differences between any pair of knowledge maps.  Intuitively, learning maps reflect the \textit{change}
in knowledge across two maps.

>>>>

We observed a similar set of results when we restricted our estimates of
participants' knowledge for questions about each lecture to consider only
their performance on other questions about the \textit{same} lecture.
Specifically, for Quiz~1, participants' knowledge of \textit{Four Fundamental
Forces}-related questions, estimated from their performance on other
\textit{Four Fundamental Forces}-related questions, was predictive of their
ability to answer those questions correctly ($OR = 15.934,\ 95\%\
\textnormal{CI} = [5.173,\ 38.005],\ \lambda_{LR} = 40.971,\ p = 0.001$). The
same was true of participants' estimated knowledge for \textit{Birth of
Stars}-related questions based on their performance on other \textit{Birth of
Stars}-related questions ($OR = 9.775,\ 95\%\ \textnormal{CI} = [2.93,\
25.08],\ \lambda_{LR} = 13.924,\ p = 0.001$). These within-lecture knowledge
estimates also predicted success on questions about both lectures when we
computed them analogously for Quiz~2 (\textit{Four Fundamental Forces}: $OR =
35.126,\ 95\%\ \textnormal{CI} = [5.113,\ 123.868],\ \lambda_{LR} = 32.251,\
p < 0.001$; \textit{Birth of Stars}: $OR = 4.717,\ 95\%\ \textnormal{CI} =
[2.021,\ 9.844],\ \lambda_{LR} = 16.788,\ p < 0.001$). For Quiz~3, we found that
within-lecture knowledge estimates predicted participants' success on
\textit{Birth of Stars}-related questions ($OR = 16.902,\ 95\%\ \textnormal{CI}
= [3.353,\ 53.265],\ \lambda_{LR} = 23.233,\ p < 0.001$) but not on \textit{Four
Fundamental Forces}-related questions ($OR = 2.485,\ 95\%\ \textnormal{CI} =
[0.724,\ 8.366],\ \lambda_{LR} = 1.984,\ p = 0.170$). This may indicate that the
within-lecture knowledge estimates are susceptible to ceiling effects in
participants' quiz performance. On Quiz~3, after viewing both lectures, no
participant answered more than three \textit{Four Fundamental Forces}-related
questions incorrectly, and all but five participants (out of 50) answered two or
fewer questions incorrectly. (This was the only subset of questions about either
lecture, across all three quizzes, for which this was true.) Because of this,
when we held out one incorrectly answered
\textit{Four Fundamental Forces}-related question from a given participant's
Quiz~3 responses and estimated their knowledge at its embedding coordinate using
the remaining \textit{Four Fundamental Forces}-related questions they answered,
for 90\% of participants, that estimate leveraged information about at most a
single other question they were \textit{not} able to correctly answer. This
homogeneity in participants' success on questions used to estimate their
knowledge may have hurt our ability to accurately characterize the specific (and
by Quiz~3, relatively few) aspects of the lecture content they did \textit{not}
know about. Taken together, these results suggest that our knowledge estimates
can reliably distinguish between questions about different content covered by a
single lecture, provided there is sufficient diversity in participants' quiz
responses to extract meaningful information about both what they know and what
they do not know.

Finally, when we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture, we observed a somewhat different pattern of results.
Here we found that before viewing either lecture (i.e., on Quiz~1),
participants' abilities to answer \textit{Four Fundamental Forces}-related
questions could not be predicted from their responses to \textit{Birth of
Stars}-related questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\
9.088],\ \lambda_{LR} = 0.712,\ p = 0.404$), nor could their abilities to
answer \textit{Birth of Stars}-related questions be predicted from their
responses to \textit{Four Fundamental Forces}-related questions ($OR = 1.522,\
95\%\ \textnormal{CI} = [0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$).
Similarly, we found that participants' performance on questions about either
lecture could not be predicted given their responses to questions about the
other lecture after viewing \textit{Four Fundamental Forces} but before viewing
\textit{Birth of Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces}
questions given \textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\
\textnormal{CI} = [0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 2.199,\ 95\%\ \textnormal{CI} = [0.711,\ 5.623],\
\lambda_{LR} = 2.304,\ p = 0.141$). Only after viewing \textit{both} lectures
(i.e., on Quiz~3) did these across-lecture knowledge estimates reliably predict
participants' success on individual quiz questions (\textit{Four Fundamental
Forces} questions given \textit{Birth of Stars} questions: $OR = 11.294,\
95\%\ \textnormal{CI} = [1.375,\ 47.744],\ \lambda_{LR} = 10.396,\ p < 0.001$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 7.302,\ 95\%\ \textnormal{CI} = [1.077,\ 44.879],\
\lambda_{LR} = 4.708,\ p = 0.038$). Taken together, these results suggest that
our knowledge estimates can be used to predict participants' success across
content areas once they have received some training on both the content about
which their knowledge is estimated and the content used to construct these
estimates.

@@

Our work makes several contributions to the study of how people acquire
conceptual knowledge. First, from a methodological standpoint, our modeling
framework provides a systematic means of mapping out and characterizing
knowledge in maps that have infinite (arbitrarily many) numbers of coordinates,
and of ``filling out'' those maps using relatively small numbers of multiple
choice quiz questions. Our experimental finding that we can use these maps to
predict responses to held-out questions has several psychological implications
as well. For example, concepts that are assigned to nearby coordinates by the
text embedding model also appear to be ``known to a similar extent'' (as
reflected by participants' responses to held-out questions;
Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. How participants' knowledge falls off with spatial distance is
captured by the knowledge maps we infer from their quiz responses
(e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words,
our study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

@@

To generate our estimates, we placed a set of 39 radial basis functions (RBFs)
throughout the embedding space, centered on the 2D projections for each
question (i.e., we included one RBF for each question). At coordinate $x$, the
value of an RBF centered on a question's coordinate $\mu$, is given by:
\begin{equation}
    \mathrm{RBF}(x, \mu, \lambda) = \exp\left\{-\frac{||x - \mu||^2}{\lambda}\right\}.
    \label{eqn:rbf}
\end{equation}
The $\lambda$ term in the RBF equation controls the ``smoothness'' of the
function, where larger values of $\lambda$ result in smoother maps. In our
implementation we used $\lambda = 50$.  Next, we estimated the ``knowledge''
at each coordinate, $x$, using:
\begin{equation}
    \hat{k}(x) = \frac{\sum_{i \in \mathrm{correct}} \mathrm{RBF}(x, q_i, \lambda)}{\sum_{j = 1}^N \mathrm{RBF}(x, q_j, \lambda)}.
    \label{eqn:rbf-knowledge}
\end{equation}
Equation~\ref{eqn:rbf-knowledge} computes the weighted proportion of correctly
answered questions, where the weights are given by how nearby (in the 2D space)
each question is to the $x$. We also defined \textit{learning maps} as the
coordinate-by-coordinate differences between any pair of knowledge maps.
Intuitively, learning maps reflect the \textit{change} in knowledge
across two maps.
